---
title: "GBIF dataset MIDS calculator"
author: "Mathias Dillen"
date: "10/14/2020"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T)
```

```{r packages,include=F}
library(tidyverse)
library(jsonlite)
library(magrittr)
```

## Set working directory

```{r setwd}
setwd("D:/apm/mids/gbifmids")
```

## Read occurrence file

Download a GBIF archive to your working directory, unzip and import the occurrence file.

```{r import}
occ = read_tsv("occurrence.txt",
               col_types = cols(.default = "c"),
               quote="")
```

## Parse informationWithheld to columns

```{r infwith, include=F}
#infwith is a function to set value to known:withheld if its DwC name is present 
#in the dwc:informationWithheld field

#behold the hideous syntax
infwith <- function(df,
                    colname) {
  df %<>%
    mutate("{colname}" := 
             ifelse(grepl(colname,
                          informationWithheld),
                    "known:withheld",
                    .data[[colname]])
    )
  return(df)
}

unksub <- function(df,
                   colname) {
  unks = c("unknown",
           "unknown:undigitized")
  df %<>%
    mutate("{colname}" := 
             ifelse(.data[[colname]]%in%unks,
                    NA,
                    .data[[colname]])
    )
  return(df)
}
```

Set the value for the property of a record to `known:withheld` if its Darwin Core property name is present in the `dwc:informationWithheld` field. Also set values of `unknown` or `unknown:undigitized` as blanks. These are considered equivalent to undigitized data, which may or may not be available. Contrast with unknown:missing or unknown:undecipherable, which indicate digitization of these data has at least been attempted.

This is currently done for the following Darwin Core properties, using a custom function:

```{r infowithheld}
usedproperties = c("locality",
                   "verbatimLocality",
                   "decimalLatitude",
                   "decimalLongitude",
                   "county",
                   "recordedBy",
                   "recordedByID",
                   "eventDate",
                   "year",
                   "verbatimEventDate")

for (i in 1:length(usedproperties)) {
  occ = infwith(occ,
                usedproperties[i])
  occ = unksub(occ,
               usedproperties[i])
}
```

## MIDS level -1

Some specimens may not even achieve MIDS level 0. This happens if they have no physical identifier or if they have no value to indicate where they are curated.

```{r minus}
occminus = occ %>%
  filter((is.na(catalogNumber)&
            is.na(otherCatalogNumbers))|
           (is.na(institutionCode)&
              is.na(collectionCode)&
              is.na(ownerInstitutionCode)&
              is.na(institutionID)&
              is.na(collectionID)&
              is.na(rightsHolder)))

occminus$mids = -1

occ %<>%
  filter(!occurrenceID%in%occminus$occurrenceID)
```

## MIDS level 0

Specimens are stuck at level 0 mainly if they have no (proper) scientific name. This can be checked by using the Taxon Name Resolution Service (http://tnrs.iplantc.org/). GBIF performs a validation step during ingestion, so we can take the output of this by removing records flagged with `TAXON_MATCH_NONE`.

```{r zero}
occ0 = occ %>%
  filter(is.na(scientificName)|
           grepl("TAXON_MATCH_NONE",
                 issue))

occ0$mids = 0

occ %<>%
  filter(!occurrenceID%in%occ0$occurrenceID)
```

## MIDS level 1

Specimens at level 1 have a proper scientific name, but are missing one of the following:

* No value for `dwc:county`, `dwc:locality`, `dwc:verbatimLocality` and decimal coordinates.
* No value for `dwc:recordedBy`. No value for `gbif:recordedByID`.
* No value for `dwc:eventDate`, `dwc:year` and `dwc:verbatimEventDate`.

```{r one}
occ1 = occ %>%
  filter(
    (is.na(county)&
       is.na(locality)&
       is.na(verbatimLocality)&
       is.na(decimalLatitude)&
       is.na(decimalLongitude))|
    (is.na(recordedBy)&
       is.na(recordedByID))|
    (is.na(eventDate)&
       is.na(year)&
       is.na(verbatimEventDate))
  )

occ1$mids = 1

occ %<>%
  filter(!occurrenceID%in%occ1$occurrenceID)
```

## MIDS level 2

Specimens are at level 2 if they have no image or if GBIF flagged them with a certain issue.

```{r two}
occ2 = occ %>%
  filter(is.na(mediaType),
         !is.na(issue))

occ2$mids = 2

occ %<>%
  filter(!occurrenceID%in%occ2$occurrenceID)
```

## MIDS level 3

Everything remaining. Summary stats:

```{r three}
occ$mids = 3

summ= tibble(n = c(
  dim(occminus)[1],
  dim(occ0)[1],
  dim(occ1)[1],
  dim(occ2)[1],
  dim(occ)[1]))

summ %<>%
  mutate(perc = round(100*n/sum(n),
                      digits=2))

print(
  paste0("Number of records at -1: ",
         dim(occminus)[1],
         " or ",
         summ$perc[1],
         "%")
)
print(
  paste0("Number of records at 0: ",
         dim(occ0)[1],
         " or ",
         summ$perc[2],
         "%")
)
print(
  paste0("Number of records at 1: ",
         dim(occ1)[1],
         " or ",
         summ$perc[3],
         "%")
)
print(
  paste0("Number of records at 2: ",
         dim(occ2)[1],
         " or ",
         summ$perc[4],
         "%")
)
print(
  paste0("Number of records at 3: ",
         dim(occ)[1],
         " or ",
         summ$perc[5],
         "%")
)

```

